Demography and selection during maize domestication

### 2015-04-30 manuscript.tex
updated

\title{Demography and selection during maize domestication}
\author{Timothy M. Beissinger\affil{1}{University of
    California,Davis}, Li Wang\affil{2}{Iowa State University}, Arun
  Durvasula\affil{1}{}, Kate Crosby\affil{1}{}, \and Jeffrey
  Ross-Ibrarra\affil{1}{}\affil{2}{Genome Center, UC
    Davis}\affil{3}{Center for population biology, UC Davis} }

@@

\subsection{Demography of maize domestication}
Text goes here

>>>>

\title{Demography and selection during maize domestication}
\author{Timothy M. Beissinger\affil{1}{University of
    California,Davis}, Li Wang\affil{2}{Iowa State University}, Arun
  Durvasula\affil{1}{}, Kate Crosby\affil{1}{}, Matthew Hufford\affil{2}{}, \and Jeffrey
  Ross-Ibrarra\affil{1}{}\affil{3}{Genome Center, UC
    Davis}\affil{4}{Center for population biology, UC Davis} }

@@

\subsection{Demography of maize domestication}
To explore whether differences in purifying selection between maize
and teosinte can be explained purely by demographic processes, we
first estimated the parameters of maize domestication using large-scale
sequencing data involving 23 inbred maize
landraces and 13 teosinte inbred lines included in the HapMap 2 panel
\cite{hapmap2} \textcolor{red}{(supplemental table here)}. The maize
lines were collected from across the Americas and the teosinte lines
came from central Mexico. Before estimating demography, we compared
the site frequency spectrum (SFS) in genic and non-genic regions, and
observed substantial differences in the evolution of these classes of
sites. For both maize and teosinte, the SFS within genes showed a
dearth of low-frequency alleles and Tajima's D
\textcolor{red}{(reference here)} was therefore shifted
to more positive values \textcolor{red}{(figure here)}. This is consistent with the aforementioned purifying
selection and indicates that genic regions are not evolving
neutrally. Therefore, for demographic modeling we restricted analysis
to non-genic sites.

We used diffusion approximation as implemented in
dadi \cite{dadi} to find the domestication parameters that best explain
the joint site frequency spectrum of maize and teosinte.  The model we optimized began with an
equilibrium ancestral population of size $N_a$
splitting into separate maize and teosinte populations $T_B$ generations in the past. Moving forward in
time from the split, teosinte maintained the ancestral population size, while
maize experienced an immediate effective population size change to
$N_b$ individuals, followed by exponential growth to size
$N_m$. Additionally, since the split a $M_{tm}$ individuals migrated
from teosinte into maize and $M_{mt}$ individuals migrated from maize
to teosinte.

The most likely model suggested an ancestral for $\theta$ ($4N\mu$) of
0.014734, which is similar to previous estimates
\textcolor{red}{(references here)}. Assuming a mutation rate of $\mu =
3 \times 10^{-8}$ \textcolor{red}{(reference here)}, this suggests an effective
population size of $N_a = 122,783$ individuals, a population split
$T_B = 15,523$ generations in the past, a maize bottleneck
effective populations size of $N_b = 6,455$ individuals (5.26\% of
$N_a$), a modern maize effective population size of $N_m = 366,973$
individuals, $M_{tm} = 1.35$ migrants per generation from teosinte to
maize, and $M_{mt} = 1.72$ migrants per generation from maize to
teosinte \textcolor{red}{(figure here)}. We note that a population
split over 15 thousand generations before present precedes estimates
from 
archaeological data which suggests maize domestication began
approximately 9,000 years before present \textcolor{red}{(reference
  here)}. This could result from multiple generations per year, or in
may reflect teosinte population structure that was present before
domestication. Also, note that the genetic time of the population
split must precede morphological changes that could be identified
morphologically.   Additionally, because recent expansion is
most evidenced by rare alleles, and because these data provide low
power to detect rare alleles, we expect that the estimate of  $N_m =
366,973$, or $\sim 3N_a$, is likely an underestimate. \textcolor{red}{(This
  is where evaluation using Kate's SFS should go)}.
